---
title: "CanopyLab field trial review"
subtitle: "Five seasons, six stations, five hybrids, three protection programmes"
author: "CanopyLab Agronomics — Trialling & Analytics"
date: last-modified
format:
  html:
    theme: brand
    toc: true
    toc-location: left
    code-fold: true
    code-summary: "Show the code"
    fig-width: 8
    fig-height: 4.5
    embed-resources: true
execute:
  warning: false
  message: false
---

::: {.callout-important appearance="simple"}
**This project contains synthetic data and analysis created for demonstration
purposes only.** CanopyLab Agronomics is a fictional crop science company. Every
station, hybrid, protection programme and yield figure below was generated by an
algorithm for this demonstration.
:::

```{r}
#| label: setup
#| include: false

library(canopytrials)
library(dplyr)
library(ggplot2)
library(gt)
library(DBI)

# The package finds the data whether this renders from the project root, from a
# scheduled job on Posit Connect, or from a CI runner.
trials <- read_trials()
sites <- read_sites()
colors <- canopylab_colors()
```

## What this report answers

Three questions, in the order the trialling team asks them:

1. **Does the protection programme pay?** And by how much, per hybrid.
2. **Where does it pay most?** Stations differ in soil and rainfall, and disease
   follows the water.
3. **What happens in a bad year?** 2023 ran about 30% below normal rainfall.

The report reads `r format(nrow(trials), big.mark = ",")` harvested plots from
`r n_distinct(trials$site_name)` stations over
`r n_distinct(trials$season)` seasons. Every number is computed at render time,
so a re-render after next harvest updates the prose as well as the charts.

## 1. Does the programme pay?

```{r}
#| label: tbl-response
#| tbl-cap: "Mean yield and the gain over each hybrid's own untreated plots"

response <- yield_response(trials)

response |>
  select(variety, treatment, plots, mean_yield_t_ha, mean_disease_index, gain_vs_check_t_ha) |>
  gt(groupname_col = "variety") |>
  fmt_number(c(mean_yield_t_ha, gain_vs_check_t_ha), decimals = 2) |>
  fmt_number(mean_disease_index, decimals = 1) |>
  cols_label(
    treatment = "Programme",
    plots = "Plots",
    mean_yield_t_ha = "Mean yield (t/ha)",
    mean_disease_index = "Disease index",
    gain_vs_check_t_ha = "Gain vs check (t/ha)"
  ) |>
  data_color(
    columns = gain_vs_check_t_ha,
    palette = c(colors[["mist"]], colors[["leaf"]])
  ) |>
  tab_source_note(md("*Synthetic data, generated for demonstration purposes only.*"))
```

The gain is always measured against **the same hybrid's own untreated plots**.
That is a deliberate choice, and it lives in the package as
`canopytrials::yield_response()` so that this report, the dashboard and any future
service cannot disagree about it: hybrids differ in genetic potential, and
comparing a treated plot to the trial-wide average would credit the programme with
genetics it had nothing to do with.

```{r}
#| label: fig-response
#| fig-cap: "The grey rule spans the gap between untreated and best-protected"

plot_yield_response(response)
```

```{r}
#| label: response-numbers
#| include: false

best <- response |>
  filter(treatment == "Programme Bravo") |>
  slice_max(gain_vs_check_t_ha, n = 1)

worst <- response |>
  filter(treatment == "Programme Bravo") |>
  slice_min(gain_vs_check_t_ha, n = 1)
```

The response is not uniform, and the spread is the interesting part.
**`r best$variety`** gains `r sprintf("%.2f", best$gain_vs_check_t_ha)` t/ha from
Programme Bravo; **`r worst$variety`** gains only
`r sprintf("%.2f", worst$gain_vs_check_t_ha)`. The hybrids that gain least are the
ones carrying their own disease tolerance — they were never losing much to disease
in the first place, so there is less for a programme to give back. Which is to say:
the value of a protection programme depends on what you planted.

## 2. Where does it pay most?

```{r}
#| label: fig-sites
#| fig-cap: "Disease pressure tracks soil and rainfall, and yield follows"

by_site <- site_summary(trials) |>
  left_join(select(sites, site_name, soil_type), by = "site_name")

ggplot(by_site, aes(x = mean_disease_index, y = mean_yield_t_ha)) +
  geom_point(aes(colour = region), size = 4) +
  geom_text(
    # Direct labels instead of a station legend: six points, six names, no lookup.
    aes(label = paste0(sub(" Station$", "", site_name), " (", tolower(soil_type), ")")),
    size = 3.2, colour = colors[["stone"]], vjust = -1.2
  ) +
  scale_colour_manual(values = unname(colors[c("leaf", "sky", "grain")])) +
  scale_x_continuous(expand = expansion(mult = 0.28)) +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.14))) +
  labs(
    title = "The heavy-soil stations carry the most disease and the least yield",
    subtitle = "Each point is one station, averaged over five seasons and all programmes",
    x = "Mean disease pressure index (0-100)",
    y = "Mean yield (t/ha)",
    caption = "Synthetic data, generated for demonstration purposes only."
  ) +
  theme_canopylab()
```

Clay Flats and High Plains sit at the bottom of the yield range for different
reasons — one holds water and grows disease, the other is high, dry and cool. A
single national recommendation would be wrong at both.

## 3. What happens in a bad year?

```{r}
#| label: fig-season
#| fig-cap: "Protected plots only. The dry season compresses the spread between hybrids."

trend <- season_trend(trials, treatment = "Programme Bravo")

ggplot(trend, aes(x = season, y = mean_yield_t_ha, colour = variety)) +
  annotate(
    "rect",
    xmin = 2022.6, xmax = 2023.4, ymin = -Inf, ymax = Inf,
    fill = colors[["mist"]]
  ) +
  annotate(
    "text",
    x = 2023, y = max(trend$mean_yield_t_ha),
    label = "2023: dry", colour = colors[["stone"]], size = 3.2, vjust = -0.4
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.4) +
  scale_colour_manual(values = unname(colors[c("leaf", "sky", "grain", "loam", "clay")])) +
  scale_x_continuous(breaks = sort(unique(trend$season))) +
  coord_cartesian(clip = "off") +
  labs(
    title = "The dry season narrows the gap between hybrids",
    subtitle = "Mean yield under Programme Bravo, by season",
    x = NULL, y = "Mean yield (t/ha)",
    caption = "Synthetic data, generated for demonstration purposes only."
  ) +
  theme_canopylab()
```

```{r}
#| label: drought-numbers
#| include: false

drought <- trials |>
  mutate(dry = season == 2023) |>
  group_by(variety, dry) |>
  summarise(mean_yield = mean(yield_t_ha), .groups = "drop") |>
  tidyr::pivot_wider(names_from = dry, values_from = mean_yield, names_prefix = "dry_") |>
  mutate(loss = dry_FALSE - dry_TRUE) |>
  arrange(loss)
```

The ranking barely moves; the *gaps* do. In the dry season
**`r drought$variety[1]`** gave up `r sprintf("%.2f", drought$loss[1])` t/ha while
**`r drought$variety[nrow(drought)]`** gave up
`r sprintf("%.2f", drought$loss[nrow(drought)])` t/ha — roughly
`r sprintf("%.1f", drought$loss[nrow(drought)] / drought$loss[1])` times as much.
A hybrid recommendation built only on average yield across seasons is silently a
bet that the next season is an average one.

## The same numbers, straight from the database

The trial results also live in a DuckDB file, which is what Positron's Connections
pane browses and what the project's `.ggsql` queries chart. The aggregation below
happens in the database, not in R:

```{r}
#| label: tbl-sql
#| tbl-cap: "Aggregated by DuckDB, formatted by gt"

connection <- connect_trials_db()

by_season <- dbGetQuery(connection, "
  SELECT season,
         treatment,
         count(*)                              AS plots,
         round(avg(yield_t_ha), 2)             AS mean_yield_t_ha,
         round(avg(disease_pressure_index), 1) AS mean_disease_index
  FROM field_trials
  GROUP BY season, treatment
  ORDER BY season, treatment
")

dbDisconnect(connection, shutdown = TRUE)

by_season |>
  gt(groupname_col = "season") |>
  cols_label(
    treatment = "Programme", plots = "Plots",
    mean_yield_t_ha = "Mean yield (t/ha)", mean_disease_index = "Disease index"
  ) |>
  tab_source_note(md("*Synthetic data, generated for demonstration purposes only.*"))
```

## How this report gets published

Nothing in this document knows where it will be rendered, which is the point.

```bash
# Publish once, from Positron or the terminal
rsconnect::deployDoc("trial-report.qmd")

# Or let CI publish it on every merge to main
# .github/workflows/deploy-connect.yml
```

On Posit Connect the report becomes a scheduled job: it re-renders itself after
each harvest upload, emails the trialling team a rendered copy, and keeps every
previous version. The R version, the package versions and the system libraries are
the ones recorded in `renv.lock` — the same ones this rendered with.

---

*This project contains synthetic data and analysis created for demonstration
purposes only. All data, insights and business scenarios were artificially
generated using AI.*
